---
title: "EPFL_3mut_combi"
author: "hjm"
date: "2024-04-24"
output: html_document
---

This file is for exploring the relative affinity data on all n=3 combinations of Omicron BA.1 mutations. WT and Omicron were also tested as a reference.

From Sai: 
1. Which residues are the most important in terms of leading to affinity losses
2. whether there are higher order (synergistic) interactions amongst residues, 
3. whether there is a difference between the antibodies in terms of being more or less susceptible to binding, and 
4. if possible to maybe learn what the possible evolutionary trajectories might be (although that is probably not possible?).
5. We may want to also think about how to use this data for an ML regression model to predict affinity?

Let's break it into parts.

```{r}
#read data
df_bam<-read.delim(file = "mutations_bamlanivimab.txt",sep = ",")
df_DBR3<-read.delim(file = "mutations_DBR3.txt",sep = ",")
df_imd<-read.delim(file = "mutations_imdevimab.txt",sep = ",")
#shunxu: 1 3 2
```

```{r}
source("RBD_tree_functions.R")
library(ggplot2)
library(dplyr)
library(tidyr)
library(stringr)
library(gridExtra)
library(car)
library(purrr)
library(igraph)
library(tidyverse)
library(scales)
library(qgraph)
```

```{r}
check_within_CI(df_DBR3)
check_within_CI(df_bam)
check_within_CI(df_imd)
```


```{r}
df_DBR3<-normalise_df(df_DBR3)
df_bam<-normalise_df(df_bam)
df_imd<-normalise_df(df_imd)
```


```{r}

normalise_result_compare(df_DBR3)
normalise_result_compare(df_bam)
normalise_result_compare(df_imd)
```



take the mean of all the replicates
```{r}

mean_affinity_by_variant<-function(df){
  
  df <- df %>%
  mutate(Mutations = str_replace_all(Mutations, ",", " "))
    
  mean_df <- df %>%
  group_by(Mutations) %>%
  summarise(Mean_Relative_Affinity = mean(Affinity_wt, na.rm = TRUE))
  merged_df <- df %>%
  left_join(mean_df, by = "Mutations")
  

  return(merged_df)
}

```

```{r}
df1<-mean_affinity_by_variant(df_bam)
df2<-mean_affinity_by_variant(df_DBR3)
df3<-mean_affinity_by_variant(df_imd)
head(df1)
df1$Antibody<-"Bamlanivimab"
df2$Antibody<-"DBR3_03"
df3$Antibody<-"Imdevimab"
```

Which residues are the most important in terms of leading to affinity losses
```{r}
#use WT as reference.
#try to calculate escape score! it is a single mutation level but accross different antibodies. use the number of escaped antibody as a representation for strength. but the difference here is the data is not binary labeled as binding/non-binding, but it is continuous. what leads to affinity loss? use affinity directly as the score. but still have to normalise by the time it appears.
#calculate the affinity decrease/increase to WT, by ratio/substract. fold change in realitive affinity. -> Affinity_wt
# first calculate within mab, then think about across mabs.
# unlist "Q493R Q498R N501Y" take the same value for every mutation in the combination
single_mutation_score <- function(df) {
    # Separate mutations into individual rows
    df_expanded <- df %>%
        mutate(Mutations = str_split(Mutations, "[, ]+")) %>%
        unnest(Mutations)

    # Calculate mean and standard deviation of fold change for each mutation
    mutation_scores <- df_expanded %>%
        group_by(Mutations) %>%
        summarise(
            Mean_Affinity_WT = mean(Mean_Relative_Affinity, na.rm = TRUE),
            SD_Affinity_WT = sd(Mean_Relative_Affinity, na.rm = TRUE),
            .groups = 'drop'  # Drop the grouping structure afterwards
        )
    mutation_scores <- mutation_scores %>%
        arrange(desc(Mean_Affinity_WT)) %>%
        mutate(Rank = row_number())

    mutation_scores$Mutations[mutation_scores$Mutations == '']<-"WT"

    return(mutation_scores)
}

```

```{r}
mut_score1 <- single_mutation_score(df1)
mut_score2 <- single_mutation_score(df2)
mut_score3 <- single_mutation_score(df3)
head(mut_score1)
```

```{r}

plot_affinity_change <- function(score_df) {
    score_df <- score_df %>%
    mutate(Mutations = gsub("E584A", "E484A", Mutations))
  
  plot <- ggplot(score_df, aes(x = reorder(Mutations, Mean_Affinity_WT), y = Mean_Affinity_WT)) +
    geom_bar(stat = "identity", fill = "#00518d") +  # Set bar color to #00518d
    labs(title = "Ranking of Mutations by Mean Affinity Change", x = "Mutations", y = "Mean Affinity Change") +
    ylim(0, 1.2) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1.1, vjust = 1.5, size = 15),
      axis.text.y = element_text(size = 10),
      panel.grid = element_blank()
    ) +
    geom_hline(yintercept = 1, linetype = "dashed", color = "red")  # Keep red dashed reference line
  return(plot)
}

```

```{r}
p1<-plot_affinity_change(mut_score1)
p2<-plot_affinity_change(mut_score2)
p3<-plot_affinity_change(mut_score3)

p<-grid.arrange(p3, p1, p2, ncol = 1)

ggsave("single_mutation_barplot.pdf", p,width = 15, height = 20)
```
how I have calculated the affinity reduction mean of single mutation. now I want to extract the same info for 2-mer mutations. this how I want to do it: for each row when there is 3 mutations in the "Mutation" column, extract every possible combination of 2 mutation and let them have r
```{r}
#try 2-mer combinations. and compare with the mean of each!
two_mutation_score<-function(df){
    mut_score <- single_mutation_score(df)
    
    df_two_mut <- df %>%
    filter(Mutations != "" & !is.na(Mutations)) %>% #!!!!!
    filter(nchar(Mutations) <18) %>%
    mutate(
    Two_Mutations = lapply(str_split(Mutations, " "), combn, 2, simplify = FALSE)) %>%
    unnest(Two_Mutations) %>%
    mutate(Two_Mutations = sapply(Two_Mutations, function(x) paste(sort(x), collapse = " "))) %>%
    group_by(Two_Mutations) %>%
    summarise(Mean_Affinity_WT_2mut = mean(Affinity_wt, na.rm = TRUE)) %>%
    ungroup() %>%
    mutate(pair_id = row_number())

 
 df_individual_scores <- df_two_mut %>%
    separate_rows(Two_Mutations, sep = " ") %>%
       #mutate(pair_id = (row_number() - 1) %/% 3 + 1) %>%
    left_join(mut_score, by = c("Two_Mutations" = "Mutations")) %>%
    group_by(pair_id) %>%
    summarise(
    Combi2 = paste(Two_Mutations, collapse=" "),
    Individual_Score = list(Mean_Affinity_WT),
    Mean_Affinity_WT_2mut = first(Mean_Affinity_WT_2mut),
    .groups = 'drop'
  )%>%
  unnest_wider(Individual_Score, names_sep = "_")

  df_individual_scores <- df_individual_scores %>%
    mutate(fold_change = Mean_Affinity_WT_2mut/pmin(Individual_Score_1 , Individual_Score_2))
   
 return(df_individual_scores)
}

```

```{r}


plot_synergestic2 <- function(df_individual_scores) {
  # Fix mutation name issue
  df_individual_scores$Combi2 <- df_individual_scores$Combi2 %>%
    str_replace_all("E584A", "E484A")
  
  # Filter data to keep only mutation pairs with fold_change ≤ 1
  df_filtered <- df_individual_scores %>% filter(fold_change <= 1)
  
  # Reorder factor levels based on fold_change
  df_filtered$Combi2 <- factor(df_filtered$Combi2, 
                               levels = df_filtered$Combi2[order(df_filtered$fold_change)])
  
  # Generate the plot
  ggplot(df_filtered, aes(x = Combi2)) +
    geom_boxplot(aes(ymin = Individual_Score_1, lower = Individual_Score_1, middle = Individual_Score_1, 
                     upper = Individual_Score_2, ymax = Individual_Score_2), 
                 stat = "identity", fill = "#a2c0f0", color=NA) +
    geom_point(aes(y = Mean_Affinity_WT_2mut), color = "darkred", size = 4) +
    labs(title = "Mutation Impact on Affinity (Fold Change <= 1)",
         x = "Mutation Combination",
         y = "Mean Affinity WT Score") +
    scale_y_continuous(limits = c(0, 1.25), breaks = seq(0, 1.25, 0.25))+
    theme(
      axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
      panel.grid.major = element_blank(),  # Removes major grid lines
      panel.grid.minor = element_blank(),  # Removes minor grid lines
      panel.border = element_blank(),  # Removes the frame
      panel.background = element_blank()  # Removes background
    ) +
    geom_hline(yintercept = 1, linetype = "dashed", color = "#d63226", size = 1)  
}


plot_synergestic2(mut2_score1)
```

```{r}
mut2_score1<-two_mutation_score(df1)
mut2_score2<-two_mutation_score(df2)
mut2_score3<-two_mutation_score(df3)

mut2_score1$antibody <- "Bamlanivimab"
mut2_score2$antibody <- "DBR3_03"
mut2_score3$antibody <- "Imdevimab"
head(mut2_score1)

```

```{r}

p1<-plot_synergestic2(mut2_score1)
p2<-plot_synergestic2(mut2_score2)
p3<-plot_synergestic2(mut2_score3)


p<-grid.arrange(p3, p1, p2, ncol = 1)
 
ggsave("synergistic2_Ime_bam_DBR3.pdf", p,width = 15, height = 20)
```

```{r}
three_mutation_score<-function(df){
  mut_score<-single_mutation_score(df)

  df_individual_scores <- df %>%
    filter(Mutations != "" & !is.na(Mutations)) %>%
    filter(nchar(Mutations) <18) %>%
    # Split the three mutations and expand into rows
    mutate(Single_Mutations = str_split(Mutations, " ")) %>%
    unnest(Single_Mutations) %>%
    # Join individual mutation scores
    left_join(mut_score, by = c("Single_Mutations"="Mutations")) %>%
    # Re-group to summarize the individual scores
    group_by(Mutations) %>%
    summarise(
      Individual_Score = list(Mean_Affinity_WT),
      Mean_Affinity_WT_3mut = first(Affinity_wt),
      .groups = 'drop'
    ) 

  df_individual_scores$Individual_Score_1<- NA
  df_individual_scores$Individual_Score_2<-NA
  df_individual_scores$Individual_Score_3<- NA
 
  for(i in 1:nrow(df_individual_scores)){
   df_individual_scores$Individual_Score_1[i]<- df_individual_scores$Individual_Score[[i]][1]
    df_individual_scores$Individual_Score_2[i]<-df_individual_scores$Individual_Score[[i]][2]
   df_individual_scores$Individual_Score_3[i]<- df_individual_scores$Individual_Score[[i]][3]
  }


  df_individual_scores <- df_individual_scores %>%
    mutate(fold_change = Mean_Affinity_WT_3mut/pmin(Individual_Score_1 , Individual_Score_2 , Individual_Score_3)) %>%
    arrange(fold_change) %>%
    mutate(Rank = row_number())

return(df_individual_scores)
}
```

```{r}
mut3_score1<-three_mutation_score(df1)
mut3_score2<-three_mutation_score(df2)
mut3_score3<-three_mutation_score(df3)
```

```{r}
plot_synergestic3 <- function(df_individual_scores) {
  antibody<-df_individual_scores$Antibody[1]
  
  df_individual_scores$Mutations <- df_individual_scores$Mutations %>%
    str_replace_all("E584A", "E484A")
  df_individual_scores$Mutations <- factor(df_individual_scores$Mutations, levels = df_individual_scores$Mutations[order(df_individual_scores$fold_change,decreasing = T)])

  df_long <- df_individual_scores %>%
    pivot_longer(
      cols = c(Individual_Score_1, Individual_Score_2, Individual_Score_3),
      names_to = "Score_Type",
      values_to = "Score_Value"
    )
  
  ggplot(df_long, aes(y = Mutations, x = Score_Value, color = Score_Type)) +
    geom_text(aes(label = "|"), hjust = 0.5, vjust = 0.5, fontface = "bold", size = 3) +
    geom_point(aes(x = Mean_Affinity_WT_3mut), color = "darkred", size = 1, position = position_dodge(width = 0.5)) +  # Red dot for combined score
    labs(title = "Individual Scores for Each Mutation Combination",
         y = "Mutation Combination",
         x = antibody) +
    xlim(0, 3) +
    theme(
      axis.text.y = element_text(angle = 0, vjust = 0.5, hjust = 1),
      panel.grid.major = element_blank(),  # Removes major grid lines
      panel.grid.minor = element_blank(),  
      panel.background = element_rect(fill = "white", colour = "black") ,# Removes minor grid lines
      legend.position = "none" 
    ) +
    scale_color_brewer(palette = "Set1", name = "Score Type") +
    scale_shape_manual(values = c(16, 17, 18))
}

```


```{r}
p1<-plot_synergestic3(mut3_score1) #Bam
p2<-plot_synergestic3(mut3_score2) #DBR3
p3<-plot_synergestic3(mut3_score3) #Imd

p<-grid.arrange(p1, p3, p2, ncol = 3)
ggsave("synergistic3_ranked.pdf", p,width = 20, height = 55,limitsize = FALSE)
```

present the top10 2mer combination and 3 mer combination based on affinity decrease comparing to the mean of individual.
```{r}
plot_fold_change<-function(mut3_score, n= 50){
  top_scores <- mut3_score %>%
  arrange(fold_change) %>%
  slice_head(n=n)

p<-plot_synergestic3(top_scores)
return(p)
}
p1<-plot_fold_change(mut3_score1)
p2<-plot_fold_change(mut3_score2)
p3<-plot_fold_change(mut3_score3)
p<-grid.arrange(p1, p2, p3, ncol = 1)
ggsave("top50mut3foldchange.pdf", p, width = 16, height = 18)
```

```{r}
plot_fold_change2<-function(mut2_score, n= 50){
  top_scores <- mut2_score %>%
  arrange(fold_change) %>%
  slice_head(n=50)

p<-plot_synergestic2(top_scores)
return(p)
}

p1<-plot_fold_change2(mut2_score1)
p2<-plot_fold_change2(mut2_score2)
p3<-plot_fold_change2(mut2_score3)
p1
p<-grid.arrange(p1, p2, p3, ncol = 1)
ggsave("top50mut2foldchange_ranked.pdf", p, width = 16, height = 16)
```

```{r}
#I think should calculate the 2-mer frequency after taking the top n of 3 mer affinity-decreasing combinations!

top_scores <- mut3_score1 %>%
arrange(fold_change) %>%
slice_head(n=50)


# Expanding three-mutation combinations into all possible two-mutation combinations
top_scores <- top_scores %>%
  mutate(Two_Mutation_Combos = map(str_split(Mutations, " "), ~ combn(.x, 2, simplify = FALSE))) %>%
  unnest(Two_Mutation_Combos) %>%
  mutate(Two_Mutation_Combos = map_chr(Two_Mutation_Combos, ~ paste(sort(.x), collapse = " ")))

# Checking how often each combination appears
two_mutation_counts <- top_scores %>%
  count(Two_Mutation_Combos, sort = TRUE)  # sort = TRUE will arrange it from most to least common


head(two_mutation_counts)
```

```{r}
#which ab is more susceptible to mutations?
#plot a violin plot for fold change distribution

# Add a column to each dataframe indicating the antibody name
mut3_score1 <- mut3_score1 %>% mutate(Antibody = "Bamlanivimab")
mut3_score2 <- mut3_score2 %>% mutate(Antibody = "DBR3")
mut3_score3 <- mut3_score3 %>% mutate(Antibody = "Imdevimab")
# Remove the row with the highest fold_change in Bamlanivimab (mut3_score1)
mut3_score1 <- mut3_score1 %>% filter(fold_change != max(fold_change, na.rm = TRUE))

# Combine all dataframes into one
combined_data <- bind_rows(mut3_score3, mut3_score1, mut3_score2) %>%
  filter(fold_change < 1) 
# Create the violin plot

g <- ggplot(combined_data, aes(x = Antibody, y = fold_change, fill = Antibody)) +
  geom_violin(trim = FALSE, scale = "width", color = "black") +  # Violin plots with black borders
  labs(title = "Fold Change Distribution by Antibody",
       x = "Antibody",
       y = "Fold Change") +
  theme_minimal() +
  scale_x_discrete(limits = c("Imdevimab", "Bamlanivimab", "DBR3_03")) + # Ensure order
  scale_y_continuous(breaks = seq(0, 1.2, by = 0.2), limits = c(-0.2, 1.4)) + 
  scale_fill_manual(values = c("Imdevimab" = "#abd7aa", 
                               "Bamlanivimab" = "#27bfbf", 
                               "DBR3_03" = "#a1bf3a")) +  # Custom colors
  theme(
    legend.position = "none",  # Remove legend
    axis.text.x = element_text(angle = 0, hjust = 0.5),
          panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(), # Rotate x-axis labels if needed
  )+
  # Add significance brackets and asterisks
    geom_segment(aes(x = 1, xend = 2, y = 1.1, yend = 1.1), color = "black") +  # Imdevimab vs. Bamlanivimab
    geom_text(aes(x = 1.5, y = 1.15, label = "*"), size = 5) +  # Add * for p = 0.0363

    geom_segment(aes(x = 2, xend = 3, y = 1.2, yend = 1.2), color = "black") +  # Bamlanivimab vs. DBR3_03
    geom_text(aes(x = 2.5, y = 1.25, label = "***"), size = 5) +  # Add *** for p < 0.0001

    geom_segment(aes(x = 1, xend = 3, y = 1.3, yend = 1.3), color = "black") +  # Imdevimab vs. DBR3_03
    geom_text(aes(x = 2, y = 1.35, label = "***"), size = 5)  # Add *** for p < 0.0001



g
ggsave("violin_foldchange.pdf", g, width = 8, height = 4)
```

```{r}
# Create the box plot
ggplot(combined_data, aes(x = Antibody, y = fold_change, fill = Antibody)) +
  geom_boxplot(outlier.color = "red", outlier.shape = 16, outlier.size = 2, color = "black") +
  labs(title = "Fold Change Distribution by Antibody (Box Plot)",
       x = "Antibody",
       y = "Fold Change") +
  theme_minimal() +
  scale_x_discrete(limits = c("Bamlanivimab", "Imdevimab", "DBR3")) +  # Ensure the desired x-axis order
  theme(
    legend.position = "none",  # Remove legend
    axis.text.x = element_text(angle = 45, hjust = 1)  # Rotate x-axis labels if needed
  )
```

```{r}
# think about comparison accross mabs.
# some mabs more suseptible? how to make this conclusion?
# evaluate on 3-mer basis
# if we are looking at the fold change then there is no need to normalise across the mabs.
#

# Merge the dataframes by 'Mutation'
combined_mut3_score <- merge(merge(mut3_score1, mut3_score2, by = "Mutations", all = TRUE), mut3_score3, by = "Mutations", all = TRUE)

# Reorder columns and rename fold_change columns
combined_mut3_score <- combined_mut3_score[, c("Mutations", "fold_change.x", "fold_change.y", "fold_change")]
colnames(combined_mut3_score) <- c("Mutation", "fold_change_bamlanivimab", "fold_change_DBR3", "fold_change_imdevimab")


long_data<- combined_mut3_score %>%
  select(Mutations, fold_change.x, fold_change.y, fold_change,
         antibody.x, antibody.y, antibody) %>%
  rename(fold_change.z = fold_change,
         antibody.z = antibody) %>%
  filter(fold_change < 1) %>%
  pivot_longer(
    cols = -Mutations,  # Exclude Mutations from the pivot
    names_to = c(".value", "set"),  # Split the column names based on pattern
    names_pattern = "(.*)\\.(.)"  # Use regex to separate the fold_change and antibody names
  ) %>%
  select(Mutations, fold_change, antibody)
  

anova_result <- aov(fold_change ~ antibody, data = long_data)
summary(anova_result)
```

```{r}

# Add a column to each dataframe indicating the antibody name
mut3_score1 <- mut3_score1 %>% mutate(Antibody = "Bamlanivimab")
mut3_score2 <- mut3_score2 %>% mutate(Antibody = "DBR3_03")
mut3_score3 <- mut3_score3 %>% mutate(Antibody = "Imdevimab")

# Combine the dataframes into one
combined_data <- bind_rows(mut3_score1, mut3_score2, mut3_score3)

# Reshape the data (optional since the combined_data is already tidy)
# This ensures the columns are Mutation, fold_change, and Antibody
long_data <- combined_data %>% 
   filter(Mean_Affinity_WT_3mut < 1) %>% 
  select(Mutations, Mean_Affinity_WT_3mut, Antibody)

# Perform ANOVA
anova_result <- aov(Mean_Affinity_WT_3mut ~ Antibody, data = long_data)

# Print ANOVA summary
summary(anova_result)

# Perform Tukey's HSD test
tukey_result <- TukeyHSD(anova_result)

# Print Tukey's test results
print(tukey_result)

# Optional: Plot Tukey HSD results
plot(tukey_result)

```

```{r}
#this test didn't assume normalisation. But gives similar results.
kruskal_result <- kruskal.test(fold_change ~ antibody, data = long_data)
print(kruskal_result)
```
```{r}
TukeyHSD(anova_result)
```

```{r}

library(multcomp)
pairwise.wilcox.test(long_data$fold_change, long_data$antibody, p.adjust.method = "BH")

```

```{r}
#ok did analysis on 3 mutation behavior of the 3 mabs and we found similarity between ly555 and regn33, the 
mut3_score1$antibody <- "bamlanivimab"
mut3_score2$antibody <- "DBR3"
mut3_score3$antibody <- "imdevimab"

# Merge the dataframes by 'Mutation'
combined_mut3_score <- merge(merge(mut3_score1, mut3_score2, by = "Mutations", all = TRUE), mut3_score3, by = "Mutations", all = TRUE)

# Reorder columns and rename fold_change columns
combined_mut3_score <- combined_mut3_score[, c("Mutations", "fold_change.x", "fold_change.y", "fold_change")]
colnames(combined_mut3_score) <- c("Mutation", "fold_change_bamlanivimab", "fold_change_DBR3", "fold_change_imdevimab")


long_data<- combined_mut3_score %>%
  select(Mutations, fold_change.x, fold_change.y, fold_change,
         antibody.x, antibody.y, antibody) %>%
  rename(fold_change.z = fold_change,
         antibody.z = antibody) %>%
  pivot_longer(
    cols = -Mutations,  # Exclude Mutations from the pivot
    names_to = c(".value", "set"),  # Split the column names based on pattern
    names_pattern = "(.*)\\.(.)"  # Use regex to separate the fold_change and antibody names
  ) %>%
  select(Mutations, fold_change, antibody)
```

```{r}
length(apply(combined_mut3_score,1,function(x) any(is.na(x)))) - sum(apply(combined_mut3_score,1,function(x) any(is.na(x))))
# number of 3-mut combinations present in every mab dataset

```
```{r}
binding_site<-read.csv("EPFLbindingsite.csv")


correct<-function(mut2_score){
    mut2_score$Combi2 <- mut2_score$Combi2%>%
    str_replace_all("E584A", "E484A")
    return(mut2_score)
}
mut2_score1<-correct(mut2_score1)
correct(mut2_score2)
correct(mut2_score3)

```

```{r}

plot_network <- function(mut2_score, binding_site) {
  

  
  # Extract the antibody name from the input dataframe
  antibody_name <- unique(mut2_score$antibody)
  if (length(antibody_name) != 1) {
    stop("Input dataframe must contain rows for only one antibody.")
  }

  # Get the binding site positions for the specified antibody
  binding_positions <- binding_site[[antibody_name]]

  mut2_score <- mut2_score %>%
    separate(Combi2, into = c("Mutation1", "Mutation2"), sep = " ", remove = FALSE) %>%
    mutate(Individual_Score_1 = as.numeric(Individual_Score_1),
           Individual_Score_2 = as.numeric(Individual_Score_2))

  scores_summary <- mut2_score %>%
    dplyr::select(Mutation1, Mutation2, Individual_Score_1, Individual_Score_2) %>%
    pivot_longer(cols = c(Mutation1, Mutation2),
                 names_to = "Mutation_Pos",
                 values_to = "Mutation") %>%
    pivot_longer(cols = c(Individual_Score_1, Individual_Score_2),
                 names_to = "Score_Pos",
                 values_to = "Score") %>%
    filter(Mutation_Pos == str_replace(Score_Pos, "Individual_Score_", "Mutation")) %>%
    group_by(Mutation) %>%
    summarize(Avg_Score = mean(Score, na.rm = TRUE)) %>%
    ungroup()

  scores_summary <- scores_summary %>%
    mutate(Position = as.numeric(str_extract(Mutation, "\\d+"))) %>%
    mutate(Category = ifelse(Position %in% binding_positions, "Binding Site", "Non-Binding Site"))

  # Create an edge list for igraph
  edges <- mut2_score %>%
    dplyr::select(Mutation1, Mutation2, fold_change)

  # Create the graph from the edge list
  g <- graph_from_data_frame(edges, directed = FALSE)

  edges_to_remove <- E(g)[E(g)$fold_change > 1]

  # Delete these edges from the graph
  g <- delete_edges(g, edges_to_remove)

  V(g)$size <- scores_summary$Avg_Score[scores_summary$Mutation %in% V(g)$name]
  V(g)$size <- (2 + (1 / V(g)$size))**2

  # Set edge width inversely proportional to fold_change
  E(g)$width <- (1 / E(g)$fold_change)**3

  # Assign categories (Binding Site or Non-Binding Site)
  V(g)$Category <- scores_summary$Category[match(V(g)$name, scores_summary$Mutation)]

  # Define colors for categories
  category_colors <- c("Binding Site" = "darkred", "Non-Binding Site" = "black")

  V(g)$color <- category_colors[V(g)$Category]
  V(g)$label.dist <- (V(g)$size * 0.2)^0.7
  V(g)$frame.color <- NA 
  
  E(g)$color <- "darkgrey"
  


  return(g)
}
category_colors <- c("Binding Site" = "darkred", "Non-Binding Site" = "black")

plot_qgraph<-function(g){
  e <- as_edgelist(g,names=FALSE)
  l <- qgraph.layout.fruchtermanreingold(e,vcount=vcount(g))
  plot(g,layout=l)
}

g1<-plot_network(mut2_score1,binding_site)
g2<-plot_network(mut2_score2,binding_site)
g3<-plot_network(mut2_score3,binding_site)

par(mfrow = c(1, 3))

plot_qgraph(g3) #im
plot_qgraph(g1) #bam
plot_qgraph(g2) #DBR3


```



```{r}
pdf("igrpah3_new.pdf", width = 13, height = 5)
par(mfrow = c(1, 3))
plot_qgraph(g1) #bam
plot_qgraph(g3) #im
plot_qgraph(g2) #Dbr3

legend("topright",               # Position of the legend
       legend = names(category_colors),  # Labels for the legend
       col = category_colors,     # Colors used in the legend
       pch = 16,                  # Type of point to use, 16 is a filled circle
       pt.cex = 1.5,              # Point size
       cex = 1.5,                 # Font size of text
       bty = "n") 
dev.off()
```
```{r}
color_palette <- colorRampPalette(c("beige", "red"))
normalized_widths <- (E(g)$width - min(E(g)$width)) / (max(E(g)$width) - min(E(g)$width))
# Apply color palette
E(g)$color <- color_palette(length(E(g)$width))[order(-normalized_widths)]
plot(g)
```

```{r}
plot(g)
legend("topright",               # Position of the legend
       legend = names(category_colors),  # Labels for the legend
       col = category_colors,     # Colors used in the legend
       pch = 16,                  # Type of point to use, 16 is a filled circle
       pt.cex = 1.5,              # Point size
       cex = 0.8,                 # Font size of text
       bty = "n") 
```

```{r}
# plot heatmap for overview of data to replace fig2A.
# align variants, 3 columns tobe 3 binders,
# color  being normalised values.
# Merge all three data frames
merged_df123 <- bind_rows(df1, df2, df3) %>%
  group_by(Antibody, Variant) %>%
  summarise(Mean_Relative_Affinity = mean(Mean_Relative_Affinity, na.rm = TRUE), .groups = "drop") %>%  # Take the mean
  pivot_wider(names_from = Antibody, values_from = Mean_Relative_Affinity)  # Convert to wide format

# Replace NAs with 0 or another suitable value (if needed)
merged_df[is.na(merged_df)] <- 0

# Convert to matrix for pheatmap (remove Variant column)
heatmap_matrix <- as.matrix(merged_df[,-1])  
rownames(heatmap_matrix) <- merged_df$Variant

# Define custom color gradient (red → beige → blue)
my_colors <- colorRampPalette(c("red", "beige", "blue"))(100)

# Generate the heatmap
pheatmap(heatmap_matrix,
         color = my_colors,
         cluster_rows = TRUE,  # Cluster similar variants
         cluster_cols = TRUE,  # Cluster antibodies
         scale = "none",       # Don't normalize data
         main = "Mean Relative Affinity Across Variants",
         fontsize = 10)  # Adjust font size if needed

```

```{r}
duplicates <- bind_rows(df1, df2, df3) %>%
  group_by(Antibody, Variant) %>%
  filter(n() > 1) %>%  # Keep only duplicates
  arrange(Antibody, Variant)

# Display duplicates
print(duplicates)
```
```{r}
normalize_variant <- function(variant) {
  if (variant %in% c("o", "o_1", "o_2")) {
    return("o")
  } else {
    return(variant)
  }
}
library(pheatmap)
merged_df123 <- bind_rows(df1, df2, df3) %>%
  mutate(Variant = sapply(Variant, normalize_variant)) %>%  # Standardize variant names
  group_by(Antibody, Variant) %>%
  summarise(Mean_Relative_Affinity = mean(Mean_Relative_Affinity, na.rm = TRUE), .groups = "drop") %>%  # Take mean for duplicates
  pivot_wider(names_from = Antibody, values_from = Mean_Relative_Affinity) 

#merged_df123[is.null(merged_df123)] <-NA
merged_df123[is.na(merged_df123)] <- -999
# Convert to matrix for pheatmap (remove Variant column)
heatmap_matrix <- as.matrix(merged_df123[,-1])  
rownames(heatmap_matrix) <- merged_df123$Variant

# Define custom color gradient (red → beige → blue)

heatmap_min <- min(heatmap_matrix[heatmap_matrix != -999])  # Smallest valid value
heatmap_max <- max(heatmap_matrix[heatmap_matrix != -999])  # Largest valid value

# **Fix: Define proper breaks only for valid values**
breaks_list <- seq(heatmap_min, heatmap_max, length.out = 101) 
breaks_list <- c(-1000, breaks_list)

my_colors <- colorRampPalette(c("darkred", "beige", "#00518d"))(100)
final_colors <- c("gray", my_colors)

desired_order <- c("Imdevimab", "Bamlanivimab", "DBR3_03")
heatmap_matrix <- heatmap_matrix[, desired_order]

# Save as PDF
pdf("pheatmap_affinity.pdf", width = 6, height = 12)  # Adjust size as needed

# Generate the heatmap
pheatmap(heatmap_matrix,
         color = final_colors,        # Use correct color mapping
         breaks = breaks_list,        # Ensure color scale remains 0-3, with gray for missing values
         cluster_rows = TRUE,         # Enable clustering for variants
         cluster_cols = FALSE,        # Keep antibody order fixed
         scale = "none",
         main = "Mean Relative Affinity Across Variants",
         fontsize = 10,
         legend = TRUE,               # Show normal legend (0 to 3, red to blue)
         show_rownames = FALSE,       # Remove row names
         angle_col = 0)                 # Make antibody names horizontal

dev.off() 

```

```{r}
# Merge data and standardize variant names
merged_df123 <- bind_rows(df1, df2, df3) %>%
  mutate(Variant = sapply(Variant, normalize_variant)) %>%  # Standardize variant names
  group_by(Antibody, Variant) %>%
  summarise(Mean_Relative_Affinity = mean(Mean_Relative_Affinity, na.rm = TRUE), .groups = "drop") %>%  # Take mean for duplicates
  pivot_wider(names_from = Antibody, values_from = Mean_Relative_Affinity) 

# Replace NA with -999 to mark missing values
merged_df123[is.na(merged_df123)] <- -999

# Convert to matrix for pheatmap (remove Variant column)
heatmap_matrix <- as.matrix(merged_df123[,-1])  
rownames(heatmap_matrix) <- merged_df123$Variant

# Ensure antibody order is: Imdevimab → Bamlanivimab → DBR3_03
desired_order <- c("Imdevimab", "Bamlanivimab", "DBR3_03")
heatmap_matrix <- heatmap_matrix[, desired_order]

# **Fix: Get min/max values ignoring -999**
heatmap_min <- min(heatmap_matrix[heatmap_matrix != -999])  # Smallest valid value
heatmap_max <- max(heatmap_matrix[heatmap_matrix != -999])  # Largest valid value

# **Fix: Define proper breaks only for valid values**
breaks_list <- seq(0, 2, length.out = 51)  # Scale from min to max
breaks_list <- c(-1000, breaks_list)  # Add -999 as a separate break

# **Fix: Ensure gray is only used for -999, not affecting other values**
my_colors <- colorRampPalette(c("darkred", "beige", "#00518d"))(50)  # Only for real values
final_colors <- c("gray", my_colors)  # Add gray as the first color

# Save as PDF
pdf("pheatmap_affinity.pdf", width = 6, height = 12)  # Adjust size as needed

# Generate the heatmap
pheatmap(heatmap_matrix,
         color = final_colors,
         breaks = breaks_list,        # Ensure color scale remains correct
         cluster_rows = TRUE,         # Enable clustering for variants
         cluster_cols = FALSE,        # Keep antibody order fixed
         scale = "none",
         main = "Mean Relative Affinity Across Variants",
         fontsize = 10,
         legend = TRUE,               # Show normal legend (0 to 3, red to blue)
         show_rownames = FALSE,       # Remove row names
         angle_col = 0,
         border_color = NA)               # Make antibody names horizontal

dev.off()  # Close the PDF device



```
```{r}
breaks_list <- seq(0, 2, length.out = 51) 
my_colors <- colorRampPalette(c("darkred", "beige", "#00518d"))(50) # Scale from min to max
# Save as PDF
pdf("pheatmap_legend.pdf", width = 6, height = 12)  # Adjust size as needed

# Generate the heatmap
pheatmap(heatmap_matrix,
         color = final_colors,
         breaks = breaks_list,        # Ensure color scale remains correct
         cluster_rows = TRUE,         # Enable clustering for variants
         cluster_cols = FALSE,        # Keep antibody order fixed
         scale = "none",
         main = "Mean Relative Affinity Across Variants",
         fontsize = 10,
         legend = TRUE,               # Show normal legend (0 to 3, red to blue)
         show_rownames = FALSE,       # Remove row names
         angle_col = 0,
         border_color = NA)               # Make antibody names horizontal

dev.off()
```
```{r figure2.1 option A}
library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)  # for arranging g1 / g2 / g3 on one page

# 0) Build the global original order: df1 -> df2 -> df3 (after normalization)
variant_order <- bind_rows(
  df1 %>% mutate(Variant = sapply(Variant, normalize_variant)),
  df2 %>% mutate(Variant = sapply(Variant, normalize_variant)),
  df3 %>% mutate(Variant = sapply(Variant, normalize_variant))
) %>%
  mutate(.row = row_number()) %>%
  arrange(.row) %>%
  distinct(Variant, .keep_all = TRUE) %>%
  pull(Variant)

# Helper to make one barplot from a single df without merging
make_plot <- function(dfi, default_title) {
  # Try to use the Antibody name as title (if the column exists & is a single value)
  title_txt <- default_title
  if ("Antibody" %in% names(dfi)) {
    ab <- unique(dfi$Antibody)
    if (length(ab) == 1 && !is.na(ab)) title_txt <- paste0(ab, " (Centered at WT=0)")
  }

  d_plot <- dfi %>%
    mutate(Variant = sapply(Variant, normalize_variant)) %>%     # normalize
    group_by(Variant) %>%
    summarise(Mean_Affinity = mean(Mean_Relative_Affinity, na.rm = TRUE), .groups = "drop") %>%  # mean of duplicates
    mutate(
      Variant = factor(Variant, levels = variant_order),         # original global order
      Mean_Affinity = Mean_Affinity - 1                          # center at WT=1
    )

  ggplot(d_plot, aes(x = Variant, y = Mean_Affinity, fill = "#00518d")) +
    geom_bar(stat = "identity", width = 0.7) +
    theme_minimal() +
    scale_fill_identity() +
    scale_y_continuous(limits = c(-2, 2)) +
    labs(
      title = title_txt,
      x = "Variant",
      y = "Mean Relative Affinity (delta from WT)"
    ) +
    theme(
      axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8),
      axis.text.y = element_text(size = 10),
      strip.text = element_text(size = 12, face = "bold"),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      legend.position = "none"
    )
}

# 1) Make the three plots (titles fall back to df1/df2/df3 if no single Antibody value)
g1 <- make_plot(df1, "df1 (Centered at WT=1)")
g2 <- make_plot(df2, "df2 (Centered at WT=1)")
g3 <- make_plot(df3, "df3 (Centered at WT=1)")

# 2) Put 3 barplots together on one page and save
combined <- g1 / g2 / g3  # stacked vertically; use (g1 | g2 | g3) for horizontal
ggsave("overview_barplots_original_order_singlepage.pdf", combined, width = 12, height = 12)


```

```{r Figure2.1 3 targets in consistent order}
merged_df123 <- bind_rows(df1, df2, df3) %>%
  mutate(Variant = sapply(Variant, normalize_variant)) %>%  # Standardize variant names
  group_by(Antibody, Variant) %>%
  summarise(Mean_Relative_Affinity = mean(Mean_Relative_Affinity, na.rm = TRUE), .groups = "drop") %>%  # Take mean for duplicates
  pivot_wider(names_from = Antibody, values_from = Mean_Relative_Affinity) 

# Prepare data: Replace NA with 0 (but don't delete them)
df_barplot <- merged_df123 %>% mutate(across(-Variant, ~replace_na(., 1)))

# Convert to matrix for clustering (remove Variant column)
affinity_matrix <- as.matrix(df_barplot[,-1])  
rownames(affinity_matrix) <- df_barplot$Variant

# Perform hierarchical clustering **only on non-NA variants**
dist_matrix <- dist(affinity_matrix)  # Compute distance matrix
clustering <- hclust(dist_matrix)  # Hierarchical clustering
ordered_variants <- clustering$labels[clustering$order]  # Ordered variant names

# Identify variants with NA values (originally)
variants_with_na <- merged_df123 %>%
  filter(if_any(-Variant, is.na)) %>%
  pull(Variant)

# Place NA-containing variants **at the end of the x-axis order**
final_variant_order <- c(setdiff(ordered_variants, variants_with_na), variants_with_na)

# Add ordered factor levels to dataframe
df_barplot <- df_barplot %>%
  mutate(Variant = factor(Variant, levels = final_variant_order))  # Apply custom order

# Convert dataframe to long format for ggplot
df_long <- df_barplot %>%
  pivot_longer(cols = -Variant, names_to = "Antibody", values_to = "Mean_Affinity") %>%
  mutate(Mean_Affinity = Mean_Affinity - 1)  # Rescale values from 0-3 → -1.5 to 1.5

# Generate the barplot
g<-ggplot(df_long, aes(x = Variant, y = Mean_Affinity, fill = "#00518d")) +
  geom_bar(stat = "identity", width = 0.7) +  # Standard bar width
  facet_wrap(~Antibody, ncol = 1, scales = "free_y") +  # Align across antibodies
  theme_minimal() +  # Clean theme
  scale_fill_identity() +  # Apply exact color (#00518d)
  scale_y_continuous(limits = c(-2, 2 )) +  # Fix y-axis scale from -1.5 to 1.5
  labs(title = "Mean Relative Affinity of Variants (Clustered, NA on Right, Rescaled)",
       x = "Variant",
       y = "Mean Relative Affinity") +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8),  # Rotate variant labels
    axis.text.y = element_text(size = 10),
    strip.text = element_text(size = 12, face = "bold"),  # Facet labels
    panel.grid.major = element_blank(),  # Remove major grid
    panel.grid.minor = element_blank(),  # Remove minor grid
    legend.position = "none"  # Remove legend (since color is fixed)
  )


ggsave("overview_barplot.pdf", g, width = 10, height = 10)

```


```{r}

# Ensure column names match antibodies correctly
 mut_score1$Antibody <- "Bamlanivimab"
 mut_score2$Antibody <- "DBR3_03"
 mut_score3$Antibody <- "Imdevimab"

# Merge all three datasets
merged_mut_df <- bind_rows(mut_score1, mut_score2, mut_score3) %>%
  select(Antibody, Mutations, Mean_Affinity_WT) %>%
  pivot_wider(names_from = Antibody, values_from = Mean_Affinity_WT)

# Convert to matrix for pheatmap (remove Mutations column)
heatmap_matrix <- as.matrix(merged_mut_df[,-1])  
rownames(heatmap_matrix) <- merged_mut_df$Mutations

# Order columns to: Imdevimab → Bamlanivimab → DBR3_03
desired_order <- c("Imdevimab", "Bamlanivimab", "DBR3_03")
heatmap_matrix <- heatmap_matrix[, desired_order]

# Define custom color gradient (red → beige → blue)
my_colors <- colorRampPalette(c("darkred", "beige", "#00518d"))(100)

# Define breaks based on min/max values (without NA)
heatmap_min <- min(heatmap_matrix, na.rm = TRUE)  
heatmap_max <- max(heatmap_matrix, na.rm = TRUE)
breaks_list <- seq(heatmap_min, heatmap_max, length.out = 101)  # Scale from min to max

# Save as PDF
pdf("pheatmap_mut_affinity.pdf", width = 4, height = 6)  # Adjust size as needed

# Generate the heatmap
pheatmap(heatmap_matrix,
         color = my_colors,          # Use correct color mapping
         breaks = breaks_list,       # Ensure color scale remains correct
         cluster_rows = FALSE,       # Keep row order (mutations)
         cluster_cols = TRUE,        # Enable column clustering
         scale = "none",
         main = "Mean Relative Affinity of Mutations",
         fontsize = 10,
         legend = TRUE,              # Show normal legend
         show_rownames = TRUE,       # Show mutation names
         angle_col = 0,              # Make antibody names horizontal
         border_color = NA)          # Remove grid lines between cells
dev.off()  # Close the PDF device

```

```{r}
library(ggplot2)
library(dplyr)
library(patchwork)
library(stringr)

plot_synergy_vs_additive <- function(df, binding_site_df, highlight_mutations) {
  ab_name <- df$antibody[1]
  
  # Get binding site for this antibody
  binding_positions <- as.character(binding_site_df[[ab_name]])
  
  df_processed <- df %>%
    mutate(
      Additive_Affinity = (Individual_Score_1 + Individual_Score_2) / 2,
      Epistatic_Affinity = Mean_Affinity_WT_2mut,
      
      # Parse mutation positions from Combi2
      Mutation_Positions = str_extract_all(Combi2, "\\d+"),
      
      # Check if any mutation is in the binding site
      Is_Binding_Site = sapply(Mutation_Positions, function(pos) any(pos %in% binding_positions)),
      
      # Identify top 2 strongest synergy (lowest fold_change)
      Is_Top2 = rank(fold_change, ties.method = "first") <= 2,
      Is_manual = Combi2 == highlight_mutations,
      # Assign dot color: red (top2), blue (binding site), black (others)
      Dot_Color = case_when(
        Is_Top2 ~ "darkorange",
        Is_manual ~ "yellow",
        Is_Binding_Site ~ "darkred",
        TRUE ~ "black"
      ),
      
      # Only label top2
      Label_Top2 = if_else(Is_Top2, Combi2, NA_character_)
    )
  
   p <- ggplot(df_processed, aes(x = Additive_Affinity, y = Epistatic_Affinity)) +
    geom_point(aes(color = Dot_Color), alpha = 0.8, size = 2) +
    scale_color_identity() +  # Use exact colors from Dot_Color
    geom_abline(slope = 1, intercept = 0, color = "black", linetype = "dashed") +
    geom_text(
    data = df_processed %>% filter(!is.na(Label_Top2), is.finite(Additive_Affinity), is.finite(Epistatic_Affinity)),
    aes(x = Additive_Affinity, y = Epistatic_Affinity, label = Label_Top2),
  vjust = -1, size = 3, color = "red"
) +
    labs(title = ab_name) +
    coord_cartesian(xlim = c(0, 1.5), ylim = c(0, 1.5), expand = FALSE)+
    theme_minimal() +
    theme(
      axis.title.x = element_blank(),
      axis.title.y = element_blank(),
      panel.grid = element_blank()
    )
   return(p)
}


```

```{r}
p1 <- plot_synergy_vs_additive(mut2_score1, binding_site,'E484A Q493R')
p2 <- plot_synergy_vs_additive(mut2_score2, binding_site)
p3 <- plot_synergy_vs_additive(mut2_score3, binding_site, 'N501Y S375F')

combined_plot <- wrap_plots(p1, p3, p2, nrow = 1) +
  plot_annotation(
    theme = theme(
      plot.margin = margin(5, 5, 5, 5),
      axis.title.x = element_text(size = 12, face = "bold"),
      axis.title.y = element_text(size = 12, face = "bold")
    ),
    caption = NULL,
    title = NULL
  ) &
  theme(
    axis.text = element_text(size = 9),
    panel.grid = element_blank()
  ) &
  labs(x = "Additive affinity", y = "Epistatic affinity")
combined_plot
ggsave("synergy_additive_annotated.pdf", combined_plot, width = 15, height = 5)

```

```{r}
p3
```

```{r}
p1
```
```{r}
#here prepare a csv file of mut3_score123 for submissioin as figure is not allowed to exceed 1 A4 paper
mut3_score123<-rbind(mut3_score1,mut3_score2,mut3_score3)
colnames(mut3_score123)
mut3_score123<-mut3_score123[,-2]
write.csv(mut3_score123,file = "mut3_score123.csv")
```

